---
title: "Template"
author: 
  - "Mark Barkalov (2782988)"
  - "Joël Dijkstra (2854215)"
  - "Shabana Esmati (2868438)"
  - "Jasper Hoogendoorn (2856970)"
  - "Giovanni Koek (2857652)"
  - "Robin Kolmus (2862402)"
  - "Elijah Kruize (2868400)"
  - "Tim van der Laan (2854528)"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages('ggplot2')
install.packages('dplyr')
install.packages('tidyr')
install.packages('readr')
install.packages('cbsodataR')
install.packages('sf')
```

```{r packages}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(tidyverse)
library(cbsodataR)
library(sf)
```

# Title Page

Include your names

Mark Barkalov (2782988)

Joël Dijkstra (2854215)

Shabana Esmati (2868438)

Jasper Hoogendoorn (2856970)

Giovanni Koek (2857652)

Robin Kolmus (2862402)

Elijah Kruize (2868400)

Tim van der Laan (2854528)

Include the tutorial group number

Group 4

Include your tutorial lecturer's name

Miss. C. Schouwenaar

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document.

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
dataset <- midwest
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1: this is a variable for the house to income ratio we used for the Netherlands, Zeeland and Flevoland.

```{r setup gen_var1, include = FALSE}
source("Gioravalav.R", echo = TRUE)
```

```{r gen_var1}
new_row <- data[5, ] / data[3,]  # Element-wise division of average Netherlands house prices divided by average personal income
new_row_fle <- data[11, ] / data[19,]  # Element-wise division of Flevoland's houseprices divided by the average income in Flevoland
new_row_zee <- data[16, ] / data[20,]  # Element-wise division of Zeeland's houseprices divided by the average income in Zeeland
```

Variable 2: This variable is the change in house prices between the years 2023 and 2019 in each province.

```{r setup gen_var2, include = FALSE}
source("Gioravalav.R", echo = TRUE)
```

```{r gen_var2}
dif <- c() # give variable dif the value of a vector

for(i in 1:nrow(data2))
{
  dif[i] = data2$"2023"[i] / data2$"2019"[i] # divides the house prices in 2023 by the houseprices in 2019 
}
```

Variable 3: This variable is the calculation of the growth of new businesses in each province

```{r setup gen_var3, include = FALSE}
source("Spatial.R", echo = TRUE)
source("Subgroup.R", echo = TRUE)
```

```{r gen_var3}
busi <- c() # give variable busi the value of a vector

for(i in 1:nrow(df_2019))
{
  busi[i] = df_2019$"Vestigingen (Aantal).x"[i] / df_2019$"Vestigingen (Aantal).y"[i] # Element division  
}
```

## 3.3 Visualize temporal variation

```{r include=FALSE}
source("Gioravalav.R", echo = TRUE)
```

```{r}
graph_data <- data.frame(
  Year = colnames(data)[1:5],  # Extract years
  Netherlands = as.numeric(data[5, 1:5]),  # Row 5: Netherlands
  Flevoland = as.numeric(data[6, 1:5]),  # Row 6: Flevoland
  Zeeland = as.numeric(data[7, 1:5])   # Row 7: Zeeland
)

# Convert data to long format
graph_data_long <- pivot_longer(graph_data, cols = -Year, names_to = "Region", values_to = "Value")

# Create the multi-line plot
ggplot(graph_data_long, aes(x = Year, y = Value, color = Region, group = Region)) + 
  geom_line(linewidth = 1) + 
  geom_point(size = 2) + 
  labs(
    title = "House Prices Relative to Average Personal Income (2019-2023)",
    x = "Year",
    y = "Relative Value"
  ) +
  theme_bw()
```

## 3.4 Visualize spatial variation

```{r visualise_map, include=FALSE}
source("Spatial.R", echo = TRUE)
```

```{r setup visualise_map}
# Create the map using ggplot2
ggplot(data2) +
  geom_sf(aes(fill = `2020`)) +  # Mapping values from the 2020 column
  scale_fill_gradient(low = "blue", high = "red", limits = c(150, 500)) +
  theme_minimal() +
  labs(title = "Map of House prices per region", fill = "House prices in 2020\nx1000 in euros")

# Create the map using ggplot2
ggplot(data2) +
  geom_sf(aes(fill = `2023`)) +  # Mapping values from the 2020 column
  scale_fill_gradient(low = "blue", high = "red", limits = c(150, 500)) +
  theme_minimal() +
  labs(title = "Map of House prices per region", fill = "House prices in 2023\nx1000 in euros")


```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

Does the growth of new businesses have a direct impact on the change in WOZ value?

```{r setup visualise_subpopulations, include=FALSE}
source("Subgroup.R", echo = TRUE)
```

```{r visualise_subpopulations}
data <- data.frame(
  Group = c("Bottom 4", "Middle 4", "Top 4"),
  Category = rep(c("Growth of New Companies", "Change in WOZ"), each = 3),
  Value = c(bottom, middle, top, bottom_woz, middle_woz, top_woz)
)

# Create the double bar plot
ggplot(data, aes(x = Group, y = Value, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Growth of New Companies and WOZ change for the 3 groups", 
       x = "Group", y = "Average Value", fill = "Category")
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r include=FALSE}
source("Gioravalav.R", echo = TRUE)
```

```{r}
ggplot(graph_data, aes(x = Year, y = Value, group = 1)) +  # Explicitly set `group = 1`
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "red", size = 2) +
  annotate("text", label = "Event from 2013 to 2022", size = 4, x = 2, y = 325) +
  annotate("rect", xmin = 1, xmax = 4, ymin = 250, ymax = 317, alpha = .2) +
  labs(
    title = "Average house prices in the Netherlands",
    x = "Year",
    y = "Value"
  ) +
  theme_bw()
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: <https://github.com/jasperdaniel06/Team-Daap>

## 5.2 Reference list

Use APA referencing throughout your document.
